---
title: "HW3"
author: "Nick Crites"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

# 1

### 1A

```{r}
library(faraway)
data(prostate)
lm1 <- lm(lpsa ~ lcavol + lweight + age + lbph + svi + lcp + gleason + pgg45, prostate)
summary(lm1)
```

Let's compute the confidence intervals associated with age at 90 and 95%:

```{r}
confint(lm1, c("age"), .95)
confint(lm1, c("age"), .90)
```

Because we see in our regression summary that age has a P-value of ~.08, when held to 95% confidence interval - or holding it to the standard of .05, we see that we cannot say age is a significant predictor because the P-value is too high. However, because the confidence interval is just below zero at 90% confidence, age is a significant predictor at that level. Therefore, we can reject the null hypothesis at a 95% confidence interval.

### 1B

```{r}
require(ellipse)
plot(ellipse(lm1, c('age', 'lbph')), type = "l")
points(0, 0, pch = 1)
abline(v= confint(lm1)['age',], lty = 2)
abline(h= confint(lm1)['lbph',], lty = 2)
```
According to the joint-null hypothesis cannot be rejected because `age = lbph = 0` lies within the confidence region. Individually, the null hypotheses`age = 0` and `lbph = 0` cannot be rejected because 0 lies within the confidence intervals.

### 1C

```{r}
nreps = 4000
set.seed(47)
tstats = numeric(nreps)
for (i in 1:nreps) {
  pros_mod2 = lm(lpsa ~ lcavol + lweight + sample(age) + lbph + svi + gleason + pgg45, prostate)
  tstats[i] = summary(pros_mod2)$coef[4,3]
}

hist(tstats)
summary(lm1)$coef[4,3]
``` 

We see that we cannot reject the null-hypothesis because the t-statistic is low - a high t-statistic in this model would be ~5. 

### 1D

```{r}
lm2 <- update(lm1, . ~ lcavol +lweight + svi)
summary(lm2)
```

```{r}
anova(lm2, lm1)
```

The ANOVA table indicates that there is not a significant improvement in fit in model 2 relative to model 1. This fact is bolstered by the residual sum of squares value being higher for model 2 than model 1. Since the P value is too high, we fail to reject the null hypothesis

# 2

### 2A

```{r}
library(faraway)
data("cheddar")
lm3 <- lm(taste ~ Acetic + H2S + Lactic, cheddar)
summary(lm3)
```

According to the regression summary's significance codes, Lactic and H2S are significant predictors at the 5% level.

### 2B

To reverse Acetic and H2S, which are on a log scale, we can use the natural log and e^x to negate this scaling:

```{r}
 lm4 <- lm(taste ~ I(exp(1)^Acetic) + I(exp(1)^H2S) + Lactic, cheddar)
summary(lm4)
```

As we see in the regression summary, only Lactic is a significant predictor at 5%. However, H2S is a significant predictor at 10%.

### 2C

We cannot use the F-test to compare these two models because `lm3` is a linear model and `lm4` is non-linear. They are also not nested, so according to page 41, we cannot use the F-test in this case. The first model, `lm3` presents a better fit of the data according to the R^2 criteria - it's value is higher. This makes sense as F-tests measure variance and the regular scale has a higher variance.

### 2D

We saw in the regression summary for (2A) that the beta coefficient for H2S was ~ 3.91. Thus, if we hold all other predictors constant and increase H2S by 0.01, we would therefore increase taste by 0.0391. 

### 2E

Our percent change can be represented as exp(x + .01) on the originial scale. Therefore, the percent change is (exp(x + 0.01) - exp(x)) / exp(x) = exp(0.01) - 1. Thus:

``` {r}
exp(1)^(0.01) -1 
```

Thus, the percent of change we observe is ~1.005%

# 3

###3A

```{r}
library(faraway)
data("teengamb")
lm5 <- lm(gamble ~ income + verbal + status + sex, teengamb)
summary(lm5)
```

According to the above regression summary, both income and sex are statistically significant at a 5% significance level. 

### 3B

The predictor `sex` is split into male (0) and female (1). These values are multiplied by the coefficient for sex (~-22.1). Thus, when all other variables are held constant, females generally spend $22.11 less than men on gambling. 

### 3C

```{r}
lm6 <- lm(gamble ~ income, teengamb)
summary(lm6)
```

```{r}
anova(lm6, lm5)
```

Because the P value is so low and a significant predictor at a 5% confidence interval, we reject the null hypothesis: that the smaller model (lm5) should be used and is a significant predictor of gamling level.s