---
title: "Lab3"
author: "Nick Crites"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

# 1

```{r}
calif = read.table("http://www.stat.cmu.edu/~cshalizi/uADA/12/hw/01/cadata.dat", header=TRUE)
dim(calif)
summary(calif) 
```


```{r}
library(scatterplot3d)
plot(calif$Longitude,calif$Latitude,pch=21, col=terrain.colors(11)[1+floor(calif$MedianHouseValue/50e3)],
bg=terrain.colors(11)[1+floor(calif$MedianHouseValue/50e3)],
xlab="Longitude",ylab="Latitude",main="Median House Prices")
legend(x="topright",legend=(50*(1:11)),fill=terrain.colors(11))
```
```{r}
plot(calif$Longitude,calif$Latitude,pch=21,
col=terrain.colors(11)[1+floor(calif$MedianHouseValue/50e3)],
bg=terrain.colors(11)[1+floor(calif$MedianHouseValue/50e3)],
cex=sqrt(calif$Population/median(calif$Population)),
xlab="Longitude",ylab="Latitude",main="Median House Prices",
sub="Circle area proportional to population")
legend(x="topright",legend=(50*(1:11)),fill=terrain.colors(11))
```

```{r}
scatterplot3d(calif$Longitude,calif$Latitude,calif$MedianHouseValue,
pch=20,scale.y=0.4,type="p",cex.symbol=0.1,
xlab="Longitude",ylab="Latitude",zlab="Median House Value")
```

```{r}
scatterplot3d(calif$Longitude,calif$Latitude,calif$MedianHouseValue,
pch=20,scale.y=0.4,type="h",cex.symbol=0.01,
xlab="Longitude",ylab="Latitude",zlab="Median House Value")
```

# 2

```{r}
fit = lm(MedianHouseValue ~ ., data=calif)
signif(coefficients(fit),3)
signif(summary(fit)$r.squared,digits=3)
```

The coefficents estimated is a description of their overall impact on `MedianHouseValue`0 our response in this model  

# 3

Based on the above coefficents, the absolute value of the largest coefficents have the greatest effect on the model. Employing this logic, the intercept most significantly effects the model output; outside of this, `Latitude` `Longitude` and `MedianIncome` have the greatest impact on model. They are all the same significance level, so we can directly compare the values.

# 4

```{r}
plot(fit)
```

The residuals in this case do not appear to have a normal distribution. In this case we would reject the null hypothesis - that our data comes from a normal distribution. We can verivfy this claim via the Shapiro-Wilkes test in number 5:

# 5

```{r}
shapiro.test(sample(resid(fit), 5000))
```

Based on the above shapiro wilkes test, we reject the null hypothesis because our p value is less than 0.05.

# 6

```{r}
fit0 = lm(log(MedianHouseValue) ~ (MedianIncome) + MedianHouseAge + TotalRooms + TotalBedrooms + Population + Households + Latitude + Longitude, calif)
signif(coefficients(fit0,3))
signif(summary(fit0)$r.squared,digits=3)
```

The scaled model seems to be a better fit as the R^2 value is higher and all the predictors are still significant. The coefficents seem to be smaller in the scaled model, so this model also is more intuitive and interpretable. 