---
title: "Class_Sept22"
author: "Nick Crites"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(faraway)
data(fat)
head(fat)
```
```{r}
lmodel <- lm(brozek ~ age + weight + height + neck + chest + abdom + hip +  thigh + ankle + biceps +  forearm + wrist, data = fat)
summary(lmodel)
```

```{r}
x = model.matrix(lmodel)
x
```
#predict the mdeian male body fat

```{r}
x0 = apply(x,2,median)
y0 =  sum(x0*coef(lmodel))
y0
```

```{r}
predict(lmodel, new=data.frame(t(x0)))
predict(lmodel,  new=data.frame(t(x0)), interval="confidence")
predict(lmodel, new=data.frame(t(x0)), interval="prediction")
```

```{r}
x1  = apply(x,2,function(x) quantile(x, 0.95))
x1
predict(lmodel, new=data.frame(t(x1)))
predict(lmodel,  new=data.frame(t(x1)), interval="confidence")
predict(lmodel, new=data.frame(t(x1)), interval="prediction")
```


Next problem


```{r}
data('airpass')
fuck <- lm(log(pass) ~ year, airpass)
plot(pass ~ year, data = airpass, type = "l", ylab = "Log Passenegers")
lines(exp(predict(fuck)) ~ year, airpass)
```

Now we are going to create our lagged data:

```{r}
lagdf = embed(log(airpass$pass), 14)
colnames(lagdf) = c("y",paste0("lag", 1:13))
lagdf = data.frame(lagdf)
lagdf
```

Create time series model:
  
```{r}
armod = lm(y ~ lag1 + lag12 + lag13 , data.frame(lagdf))
summary(armod)
```

Now lets overlay instead of straight line regression:

```{r}
plot(pass~year, airpass, type = "l")
lines(airpass$year[14:144], exp(predict(armod)), lty = 2, col= 2)
#The last value  in our dataset
lagdf[nrow(lagdf),]
#predict our next value
predict(armod, data.frame(lag1=6.0684, lag12=6.033086, lag13 = 5.891644), interval = "prediction")
```

```{r}
sim_1 = function(sample_size = 500){
  x = runif(n=sample_size)*5
  y =  3 + 5*x +  rnorm(n=sample_size, mean=0, sd=1)
  data.frame(x,y)
}

set.seed(42)
sim_data_1 = sim_1()
head(sim_data_1)
```
}

```{r}
sim_2 = function(sample_size = 500){
  x = runif(n=sample_size)*5
  y =  3 + 5*x +  rnorm(n=sample_size, mean=0, sd=x)
  data.frame(x,y)
}
```

```{r}
sim_3 = function(sample_size = 500){
  x = runif(n=sample_size)*5
  y =  3 + 5*x^2 +  rnorm(n=sample_size, mean=0, sd=5)
  data.frame(x,y)
}
```

```{r}
set.seed(42)
sim_data_1 = sim_1()
sim_data_2 = sim_2()
sim_data_3 = sim_3()
plot(y~x, data = sim_1(), col = "grey", pch =20, main = "DataFrame1")
fit_1 = lm(y~x,data = sim_data_1)
abline(fit_1, col = "darkorange", lwd=3) 
#satisfies linearity and constant variance
plot(fitted(fit_1), resid(fit_1), col = "grey", pch = 20, xlab = "fitted", ylab = "residuals", main = "Data From Model 2")
abline(h=0, col="darkorange")
```
```{r}
plot(y~x, data = sim_2(), col = "grey", pch =20, main = "DataFrame2")
fit_2 = lm(y~x,data = sim_data_2)
abline(fit_2, col = "red", lwd=3)
#satisfies linearity
plot(fitted(fit_2), resid(fit_2), col = "grey", pch = 20, xlab = "fitted", ylab = "residuals", main = "Data From Model 2")
abline(h=0, col="red")
```

```{r}
plot(y~x, data = sim_3(), col = "grey", pch =20, main = "DataFrame3")
fit_3 = lm(y~x,data = sim_data_3)
abline(fit_3, col = "green", lwd=3)
#satisfies linearity
plot(fitted(fit_3), resid(fit_3), col = "grey", pch = 20, xlab = "fitted", ylab = "residuals", main = "Data From Model 2")
abline(h=0, col="green")
```

```{r}
par(mfrow = c(1,3))
hist(resid(fit_1),
     xlab = "residuals",
     ylab = "Histogram of Residuals, fit_1",
     col  = "darkorange",
     border = "dodgerblue",
     breaks = 20)
```

```{r}
par(mfrow = c(1,3))
hist(resid(fit_2),
     xlab = "residuals",
     ylab = "Histogram of Residuals, fit_2",
     col  = "darkorange",
     border = "dodgerblue",
     breaks = 20)
```

```{r}
par(mfrow = c(1,3))
hist(resid(fit_3),
     xlab = "residuals",
     ylab = "Histogram of Residuals, fit_3",
     col  = "darkorange",
     border = "dodgerblue",
     breaks = 20)
```

Lets use QQ plot and Shapiro-Wilkes test

```{r}
qqnorm(resid(fit_1), main = "Normal Q-Q plot, Fit_1", col = "darkgrey")
qqline(resid(fit_1), col = "dodgerblue", lwd=2)
```
Points should come from a straiight line, this would suggest they don't come from a normal distribution

```{r}
qqnorm(rnorm(10))
qqline(rnorm(10))
qqnorm(rnorm(25))
qqline(rnorm(25))
qqnorm(rnorm(100))
qqline(rnorm(100))
qqnorm(rnorm(1000))
qqline(rnorm(1000))
qqnorm(rnorm(100000))
qqline(rnorm(100000))
```
```{r}
qqnorm(rexp(10))
qqline(rexp(10))
qqnorm(rexp(100))
qqline(rexp(100))
qqnorm(rexp(1000))
qqline(rexp(1000))
par(mfrow = c(3,1))
```

QQ plots for model 2 and 3

```{r}
qqnorm(resid(fit_2), main = "Normal QQ plot, Fit_2", col = "darkgrey")
qqline(resid(fit_2), col = "dodgerblue", lwd = 2)

qqnorm(resid(fit_3), main = "Normal QQ plot, Fit_3", col = "darkgrey")
qqline(resid(fit_3), col = "dodgerblue", lwd = 2)
```

Shapiro Wilks Test!

```{r}
shapiro.test(resid(fit_1))
```

```{r}
shapiro.test(resid(fit_2))
```
```{r}
shapiro.test(resid(fit_3))
```
Bigger p - value means higher likelihood  of normality

